Machine Learning for Predictive Auto-Tuning with Boosted Regression Trees

### 2011-12-14 paper/draft1.tex
Aspell pass

In this paper, we propose to \emph{learn the model} using non-linear regression
modelling techniques instead of constructing a model manually. By learning the model, one can
hope to achieve elements of the best of both approaches: the search speed of model-based auto-tuning with the broad applicability
and ease of implementation of empirical auto-tuning.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=\linewidth]{illustrator/fig_allstars_mixup_580_R1.pdf}
\caption{Different arguments call for different kernels:
{\em left} is the average argument-specific empirical
auto-tuned performance across 100 random argument configurations,
{\em middle} is the average speed of our reference implementation,
{\em right} is the average speed of kernels auto-tuned for different argument
configurations than the one being tested.  For 38/100 random confurations, the
kernel auto-tuned for another problem could not even run on the GTX 295
hardware.  These points contributed a speed of 0, bringing down the average
much lower than the reference.  Good performance across a variety of inputs
requires {\em input-dependent auto-tuning}.
}
\label{fig:allstars}
\end{figure}

\begin{figure}[!t]
\centering
\includegraphics[width=\linewidth]{illustrator/fig_ntrain_munctional0_580.pdf} % on vader
\caption{The effect of the invalid configuration score ($\zeta$) and training set
size on predictive auto-tuning.
Models trained on as few as 25 configurations outperformed the reference.
Models trained on 50 or more configurations rivalled empirical auto-tuning.
Moderate values of $\zeta$ between $\log(0.25)$ and $\log(0.75)$ are best.
}
\label{fig:fig_ntrain}
\end{figure}

@@

% say what we said
In this paper, we've demonstrated a boosted regression tree-based auto-tuning
method, wherein empirical performance data is used to train a machine learning
model of performance for an instrumented GPU kernel.  In contrast to
traditional model-based auto-tuning, where an explicit model of performance is
built on the basis of an understanding of hardware inner workings, and
empirical auto-tuning, where an exhaustive set of implementation configurations
are tried, the present approach generates, from scratch, a model of kernel
performance on the basis of timing data from a user-definable number of kernel
evaluations.  This approach allows significant flexibility to navigate
tradeoffs between offline and run-time costs, and final auto-tuning
performance.  Importantly, this method treats kernels as black-boxes, allowing
the user to auto-tune in the absence of deep knowledge of hardware details
(which may even been unknowable, in the case of hardware not available at the
time of kernel creation).  This approach also frees auto-tuning performance
from a strong dependency on the accuracy (or inaccuracy) of a pre-defined
analytical model.

An important use case for the tools described here is in the development of
user-facing numerical libraries.  Such libraries are a critical component of
scientific computing infrastructure, since they abstract away implementation
details and make algorithms available to a much wider audience.  However, the
abstraction provided by libraries represents a double-edged sword: one hand,
\emph{using} the library is easier, because it presents a unified abstraction
of related functionality.  However, at the same time, any given library routine
might represent a wide range of substantially different problem configurations,
each with distinct computational issues and bottlenecks.  Auto-tuning has long
provided a solution that finesses these two issues, providing multiple
implementations under the hood, for multiple problem settings, and then using
heuristics or explicit, hand-crafted models to select the appropriate
implementation for a given set of inputs.  The development of such auto-tuned
libraries, while extremely successful, is also very difficult.  The
machine-learning-based techniques described here provide a middle ground, where
a library implementor can simply create an instrumented kernel, and allow
generic tools to automatically generate appropriate auto-tuned implementations,
with small (and controllable) run-time costs.

>>>>

In this paper, we propose to \emph{learn the model} using non-linear regression
modeling techniques instead of constructing a model manually. By learning the model, one can
hope to achieve elements of the best of both approaches: the search speed of model-based auto-tuning with the broad applicability
and ease of implementation of empirical auto-tuning.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=\linewidth]{illustrator/fig_allstars_mixup_580_R1.pdf}
\caption{Different arguments call for different kernels:
{\em left} is the average argument-specific empirical
auto-tuned performance across 100 random argument configurations,
{\em middle} is the average speed of our reference implementation,
{\em right} is the average speed of kernels auto-tuned for different argument
configurations than the one being tested.  For 38/100 random configurations, the
kernel auto-tuned for another problem could not even run on the GTX 295
hardware.  These points contributed a speed of 0, bringing down the average
much lower than the reference.  Good performance across a variety of inputs
requires {\em input-dependent auto-tuning}.
}
\label{fig:allstars}
\end{figure}

\begin{figure}[!t]
\centering
\includegraphics[width=\linewidth]{illustrator/fig_ntrain_munctional0_580.pdf} % on vader
\caption{The effect of the invalid configuration score ($\zeta$) and training set
size on predictive auto-tuning.
Models trained on as few as 25 configurations outperformed the reference.
Models trained on 50 or more configurations rivaled empirical auto-tuning.
Moderate values of $\zeta$ between $\log(0.25)$ and $\log(0.75)$ are best.
}
\label{fig:fig_ntrain}
\end{figure}

@@

% say what we said
In this paper, we've demonstrated a boosted regression tree-based auto-tuning
method, wherein empirical performance data is used to train a machine learning
model of performance for an instrumented GPU kernel.  In contrast to
traditional model-based auto-tuning, where an explicit model of performance is
built on the basis of an understanding of hardware inner workings, and
empirical auto-tuning, where an exhaustive set of implementation configurations
are tried, the present approach generates, from scratch, a model of kernel
performance on the basis of timing data from a user-definable number of kernel
evaluations.  This approach allows significant flexibility to navigate
trade-offs between offline and run-time costs, and final auto-tuning
performance.  Importantly, this method treats kernels as black-boxes, allowing
the user to auto-tune in the absence of deep knowledge of hardware details
(which may even been unknowable, in the case of hardware not available at the
time of kernel creation).  This approach also frees auto-tuning performance
from a strong dependency on the accuracy (or inaccuracy) of a pre-defined
analytical model.

An important use case for the tools described here is in the development of
user-facing numerical libraries.  Such libraries are a critical component of
scientific computing infrastructure, since they abstract away implementation
details and make algorithms available to a much wider audience.  However, the
abstraction provided by libraries represents a double-edged sword: one hand,
\emph{using} the library is easier, because it presents a unified abstraction
of related functionality.  However, at the same time, any given library routine
might represent a wide range of substantially different problem configurations,
each with distinct computational issues and bottlenecks.  Auto-tuning has long
provided a solution that finesses these two issues, providing multiple
implementations under the hood, for multiple problem settings, and then using
heuristics or explicit, hand-crafted models to select the appropriate
implementation for a given set of inputs.  The development of such auto-tuned
libraries, while extremely successful, is also very difficult.  The
machine-learning-based techniques described here provide a middle ground, where
a library developer can simply create an instrumented kernel, and allow
generic tools to automatically generate appropriate auto-tuned implementations,
with small (and controllable) run-time costs.
